---
title: "Figure 3 and S3"
output: html_notebook
---

# Figure 3

## cell type age plots

### read in cell type ages and make dotplots

```{r}

df.ct = fread('data/cell-type-ages.txt')

# dotplot
ggplot(df.ct %>% filter(total_cells > cell_count_thresh), aes(x = reorder(cell_type, rel), y = mean.time, fill = temp)) + 
  geom_hline(yintercept = median(df.sum[df.sum$temp == '28C',]$mean.time)) +
  geom_hline(yintercept = median(df.sum[df.sum$temp == '28C',]$mean.time) + sd(df.sum[df.sum$temp == '28C',]$mean.time),
             lty = 'longdash', col = 'grey') +
  geom_hline(yintercept = median(df.sum[df.sum$temp == '28C',]$mean.time) - sd(df.sum[df.sum$temp == '28C',]$mean.time),
             lty = 'longdash', col = 'grey') +
  scale_fill_manual(values = c('darkgrey', '#f77168', 'darkred')) +
  stat_summary(fun.data="mean_sdl", mult=1, 
                 geom="crossbar", width=0.5) +
  geom_point(size = 1, color='black', pch = 21, position = position_jitter(h = 0.1, w = 0, seed = 2021)) +
  facet_wrap(temp ~ ., ncol = 1) +
  coord_flip() +
  theme(text = element_text(size=6),
        legend.position = 'none') +
  monocle3:::monocle_theme_opts()


```



## cell type age model

### read in covariates df (has expression levels of signatures at basal (28C) condition) along with relative progression


```{r}

covariates_df = fread('/Users/dorrity/Dropbox/q_lab/projects/40_fish/HF4/hotfish-github/data/for-zscape/covariates-df-relative-progression.txt')
covariates_df$log_rel_progression = log2(covariates_df$rel_28) # response is log relative progression of 32/34C over 28C

```


### model

```{r}

fit.null = VGAM::vglm('log_rel_progression ~ 1',
                      data = covariates_df,
                      family = gaussianff(zero = NULL),
                      trace = FALSE)

fit = VGAM::vglm('log_rel_progression ~
                 timepoint_fix +
                 upr_index +
                 translation_index +
                 proliferation_index +
                 apoptosis_index +
                 degr_misfolded_index +
                 stress_granule_index +
                 ubq_index +
                 heat_stress_index +
                 aerobic_resp_index +
                 anaerobic_resp_index +
                 tca_cycle_index',
                      data = covariates_df,
                      family = gaussianff(zero = NULL),
                      trace = FALSE)

lr.obj <- lrtest(fit, fit.null)
test.stat <- lr.obj@Body$Chisq[2]
p.value <- lr.obj@Body$`Pr(>Chisq)`[2]
pseudo_R2 <- 1 - (logLik(fit) / logLik(fit.null))
pseudo_R2

#covariates_df$prediction = predict(fit, covariates_df)[,1][match(rownames(covariates_df), names(predict(fit, covariates_df)[,1]))]

# pull out mean coefficients
selected_coefs = coef(fit)
selected_coefs = selected_coefs[grep('\\:1', names(selected_coefs))]
tidy_cfs = lapply(selected_coefs, function(cfs){
  broom::tidy(cfs)
})

ct_names = names(selected_coefs)
tidy_cfs = do.call(rbind, tidy_cfs)
tidy_cfs$term = ct_names
row.names(tidy_cfs) = NULL
colnames(tidy_cfs) = c('value', 'term')

```

### plot model outputs

```{r}



# plot top coefficients as barplot
ggplot(tidy_cfs %>% filter(term != '(Intercept):1')) +
  geom_bar(aes(x = reorder(term, - value), y = value , fill = value), stat = 'identity', col = 'black') +
  scale_fill_gradient2(low = 'steelblue', mid = 'white', high = 'tomato') +
  monocle3:::monocle_theme_opts() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


# plot UPR against progression
ggplot(covariates_df %>% filter(upr_index < 0.05)) + # removed hatching gland to see the rest of points
  geom_point(aes(x = upr_index, y = rel_28, fill = temp.x), stroke = 0.75, pch = 21, cex = 3) +
  geom_smooth(aes(x = upr_index, y = rel_28), method = 'lm', lwd = 1, se = FALSE, col = 'grey') +
  scale_fill_manual(values = c('#f77168', 'darkred')) +
  xlab('UPR index') + ylab('relative cell type progression') +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none') +
  theme(aspect.ratio=1)

# same plot but for proliferation
ggplot(covariates_df) +
  geom_point(aes(x = proliferation_index, y = rel_28, fill = temp.x), stroke = 1, pch = 21, cex = 2.2) +
  geom_smooth(aes(x = proliferation_index, y = rel_28), method = 'lm', lwd = 1, se = FALSE, col = 'grey') +
  scale_fill_manual(values = c('#f77168', 'darkred')) +
  xlab('proliferation index') + ylab('relative cell type progression') +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none') +
  theme(aspect.ratio=1) 



```


## cell type stage deviation

```{r}

df.ct = fread('data/cell-type-ages.txt')

# show deviation as histograms/density plots
ggplot() + 
  geom_density(data = df.sum %>%
               filter(total_cells > cell_count_thresh,
                sd.time < 5),
                aes(x = deviation, fill = temp),
               lwd = 1.25, col = 'black', alpha = 0.5) +
  scale_fill_manual(values = c('darkgrey', '#f77168', 'darkred')) +
  xlim(-20,20) +
  xlab('deviation from whole-embryo expectation') +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none')


```



## age covariance

### read in coldata
```{r}

coldat = fread('data/zscape/hotfish-ref_18-48hpf_coldata.csv')

```

### compute correlations, produce heatmaps

```{r}

df = coldat %>% 
  filter(cell_type_broad != '') %>% 
  group_by(embryo, cell_type_broad, timepoint, temp) %>% 
  dplyr::summarize(mean_nn_time = mean(mean_nn_time, na.rm=TRUE))

timepoint_of_interest = 30

df.wider.28 = df %>% filter(temp == '28C', timepoint == timepoint_of_interest) %>% 
    pivot_wider(names_from = cell_type_broad, values_from = mean_nn_time, values_fill = list(mean_nn_time = 0)) %>%
    as.data.frame()
rownames(df.wider.28) = df.wider.28$embryo

df.wider.32 = df %>% filter(temp == '32C', timepoint == timepoint_of_interest) %>% 
    pivot_wider(names_from = cell_type_broad, values_from = mean_nn_time, values_fill = list(mean_nn_time = 0)) %>%
    as.data.frame()
rownames(df.wider.32) = df.wider.32$embryo

df.wider.34 = df %>% filter(temp == '34C', timepoint == timepoint_of_interest) %>% 
    pivot_wider(names_from = cell_type_broad, values_from = mean_nn_time, values_fill = list(mean_nn_time = 0)) %>%
    as.data.frame()
rownames(df.wider.34) = df.wider.34$embryo

mat.28 = df.wider.28 %>% select(-embryo, -timepoint, -temp) %>% as.matrix()
mat.32 = df.wider.32 %>% select(-embryo, -timepoint, -temp) %>% as.matrix()
mat.34 = df.wider.34 %>% select(-embryo, -timepoint, -temp) %>% as.matrix()

# filter cell types without transcriptional age values from at least 100 embryos
col_sparsity_thresh = 100
mat.28 = mat.28[,colSums(mat.28) > col_sparsity_thresh]
mat.32 = mat.32[,colSums(mat.32) > col_sparsity_thresh]
mat.34 = mat.34[,colSums(mat.34) > col_sparsity_thresh]

# compute covariances
cov.28 = cor(mat.28)
cov.32 = cor(mat.32)
cov.34 = cor(mat.34)

# set NAs to 0
cov.28[is.na(cov.28)] <- 0
cov.32[is.na(cov.32)] <- 0
cov.34[is.na(cov.34)] <- 0

# subset to cell types with values at all temperatures
cts = intersect(rownames(cov.28), rownames(cov.32))
cts = intersect(cts, rownames(cov.34))
cov.28 = cov.28[cts,cts]
cov.32 = cov.32[cts,cts]
cov.34 = cov.34[cts,cts]

# set colors
paletteLength = 12
myColor <- colorRampPalette(c('steelblue','white', "darkred"))(paletteLength)
myBreaks <- c(seq(min(cov.28,na.rm=TRUE), 0, length.out=ceiling(paletteLength/2) + 1),
              seq(max(cov.28,na.rm=TRUE)/paletteLength, max(cov.28,na.rm=TRUE), length.out=floor(paletteLength/2)))

# set ordering using the 28C heatmap
out = pheatmap(cov.28, col = myColor, breaks = myBreaks, cluster_rows = TRUE, cluster_cols = TRUE,
         show_rownames = FALSE, show_colnames = FALSE)
row.ord = out$tree_row[["order"]]
col.ord = out$tree_col[["order"]]
mat.28.reord = cov.28[row.ord, col.ord]
mat.32.reord = cov.32[row.ord, col.ord]
mat.34.reord = cov.34[row.ord, col.ord]

pheatmap(mat.28.reord, col = myColor, breaks = myBreaks, cluster_rows = FALSE, cluster_cols = FALSE,
         show_rownames = FALSE, show_colnames = FALSE)
pheatmap(mat.32.reord, col = myColor, breaks = myBreaks, cluster_rows = FALSE, cluster_cols = FALSE,
         show_rownames = FALSE, show_colnames = FALSE)
pheatmap(mat.34.reord, col = myColor, breaks = myBreaks, cluster_rows = FALSE, cluster_cols = FALSE,
         show_rownames = FALSE, show_colnames = FALSE)

# density plot of correlations for each temp
plot(density(cov.34), xlim = c(-1,1), lwd = 3, bty = 'n',col = '#D14334', ylim = c(0,4))
lines(density(cov.32), lwd = 3, col = '#E5786B')
lines(density(cov.28), lwd = 3, col = '#9F9F9F')


```



## Figure S3


## heat shock against progression

# plot heat shock against progression

```{r}


covariates_df = fread('data/covariates-df-relative-progression.txt')

ggplot(covariates_df %>% filter(!is.na(rel_28))) + 
  geom_point(aes(x = heat_stress_index, y = rel_28 %>% log2(), col = temp.x)) +
  geom_smooth(aes(x = heat_stress_index, y = rel_28 %>% log2(), col = temp.x),
              method = 'lm') +
  xlab('heat stress signature') + ylab('log2 relative progression over 28C') +
  monocle3:::monocle_theme_opts() +
   scale_color_manual(values = c('#fa8778', '#eb4934')) +
  labs(col = 'temperature')

# show as an ordered barplot by progression
ggplot(covariates_df  %>% filter(!is.na(rel_28),
                            temp.x == '34C')) + 
  geom_bar(aes(x = reorder(cell_type_broad, rel_28), y = heat_stress_index, fill = temp.x),
           stat = 'identity', position = 'dodge') +
  xlab('cell types ordered by relative progression') + ylab('heat stress signature') +
  monocle3:::monocle_theme_opts() +
  scale_fill_manual(values = c('#eb4934')) +
  labs(fill = 'temperature') +
  coord_flip() + 
  facet_grid(.~timepoint_fix)



```


## boxplots of cell type-specific acceleration

```{r}

# do a set of types
types_of_interest = c('endothelium (dorsal aorta)',
                      'notochord',
                      'neural crest + pigment cell progenitor',
                      'basal cell')

p <- ggplot(df %>% filter(cell_type %in% types_of_interest,
                          timepoint %in% c(24,30,36)), aes(x = temp, y = mean_nn_time, col = temp)) +
  geom_jitter() + 
  geom_boxplot(size = 2) +
  scale_color_manual(values = c('grey', '#fa8778', '#eb4934')) +
  ylim(15,70) + 
   theme(axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x=element_blank(),
    legend.position = "none",
    #rect = element_rect(fill = "transparent"),
    panel.background = element_rect(fill = "transparent",colour = NA),
    plot.background = element_rect(fill = "transparent",colour = NA)) +
  facet_grid(.~cell_type)

p


```

















