---
title: "Figure 4 + S4"
output: html_notebook
---

## read in packages

```{r}


plot_umap_gene_expr <- function(cds, marker, cell_size = 1, stroke_size = 0.0){
                           
  cds_subset <- cds[rowData(cds)$gene_short_name == marker,]
  marker_expr <- Matrix::t(assay(cds_subset))/colData(cds_subset)$Size_Factor
  
  meta_df <- as.data.frame(colData(cds_subset))
  meta_df$umap1 <- reducedDim(x = cds, type = "UMAP")[,1]
  meta_df$umap2 <- reducedDim(x = cds, type = "UMAP")[,2]
  
  meta_df$marker <- as.numeric(marker_expr[,1])
  meta_df$marker = with(meta_df, ifelse(marker >= 0.01, marker, NA))
  
  ggplot(meta_df, aes(x = umap1, y = umap2, color = log10(marker + 0.1))) +
    geom_point(size = cell_size, stroke = stroke_size, aes(alpha = ifelse(!is.na(marker), "1", "2"))) + 
    scale_color_viridis(option = "viridis",
          name = "log10(",marker," + 0.1)", na.value = "grey80") +
    scale_alpha_manual(values = c(1, .3)) + guides(alpha = FALSE) +
    monocle3:::monocle_theme_opts() +
    theme_void()
}

```


## translation - upr scatterplot

```{r}

# sig_df = fread('/Users/dorrity/Dropbox/q_lab/projects/40_fish/HF4/hotfish-github/data/for-zscape/hotfish-ref-signature-scores.txt')

sig_df = fread('data/zscape/hotfish-ref-signature-scores.txt')

# make embryo celltype df
df = sig_df %>% 
  filter(temp %in% c('28C','32C', '34C')) %>% 
  filter(experiment == 'HF4') %>% 
  group_by(cell_type_broad, oligo_tag, pseudostage, temp) %>% 
  summarize(mean_translation = mean(translation_index, na.rm=TRUE),
            mean_upr = mean(upr_index, na.rm=TRUE),
            mean_degr = mean(degr_misfolded_index, na.rm=TRUE),
            er_stress_index = mean(er_stress_index, na.rm=TRUE),
            mean_age = mean(mean_nn_time, na.rm=TRUE))

# scatterplot of UPR signature vs translation signature
ggplot(df %>% filter(cell_type_broad != 'notochord')) +
  geom_vline(xintercept = 0, col = '#4a4a4a', lwd = 0.5, lty = 'longdash') +
  geom_hline(yintercept = 0, col = '#4a4a4a', lwd = 0.5, lty = 'longdash') +
  geom_point(aes(x = mean_translation, y = mean_upr),
             size = 1, stroke = 0.1, shape = 20, col = 'darkgrey') +
  # highlight notochord cells
  geom_smooth(data = df %>% filter(cell_type_broad == 'notochord'),
             aes(x = mean_translation, y = mean_upr), col = 'black', method = 'lm', lwd = 0.5, se = FALSE) +
  geom_point(data = df %>% filter(cell_type_broad == 'notochord'),
             aes(x = mean_translation, y = mean_upr),
             size = 1, stroke = 0.1, shape = 20, col = 'black') +
  ylab('mean UPR per embryo-celltype') +
  xlab('mean translation per embryo-celltype') +
  ylim(-0.15,0.3) +
  xlim(-0.15,0.3) + 
  monocle3:::monocle_theme_opts()


```

## translation - upr circular plot


```{r}


sig_df = fread('data/zscape/hotfish-ref-signature-scores.txt')

# make embryo celltype df
df = sig_df %>% 
  filter(temp %in% c('28C','32C', '34C')) %>% 
  group_by(cell_type_broad, oligo_tag, pseudostage, temp) %>% 
  summarize(mean_translation = mean(translation_index, na.rm=TRUE),
            mean_upr = mean(upr_index, na.rm=TRUE),
            mean_degr = mean(degr_misfolded_index, na.rm=TRUE),
            er_stress_index = mean(er_stress_index, na.rm=TRUE),
            mean_age = mean(mean_nn_time, na.rm=TRUE))


# count up positive cells, look at fraction/percent expressed
df <- df %>%
  mutate(
    both = ifelse(mean_upr > 0 & mean_translation > 0, 1, 0),
    none = ifelse(mean_upr < 0 & mean_translation < 0, 1, 0),
    upr = ifelse(mean_upr > 0 & mean_translation < 0, 1, 0),
    tln = ifelse(mean_upr < 0 & mean_translation > 0, 1, 0),
  )

df.pct = df %>% select(cell_type_broad, temp, both, upr, tln, none) %>%
  group_by(cell_type_broad, temp) %>% 
  summarize(both = sum(both),
            upr = sum(upr),
            tln = sum(tln),
            none = sum(none),
            total = n(),
            pct_both = (both/total) * 100)

# merge in pct_both signatures from the two temps
df.pct.32 = df.pct[df.pct$temp == '32C',]
df.pct.34 = df.pct[df.pct$temp == '34C',]
df.pct$pct_both_32 = df.pct.32$pct_both[match(df.pct$cell_type_broad, df.pct.32$cell_type_broad)]
df.pct$pct_both_34 = df.pct.34$pct_both[match(df.pct$cell_type_broad, df.pct.34$cell_type_broad)]
df.pct$pct_both_temp = apply(df.pct[,c('pct_both_32','pct_both_34')], 1, mean) # mean pct_both over temperatures
df.pct$pct_both_ratio = df.pct$pct_both_temp - df.pct$pct_both # compute percent increase in fraction expressed

# circular plot
ggplot() +
  geom_hline(yintercept = 0, col = 'black', lwd = 1) +
  geom_hline(yintercept = 25, col = 'black', lwd = 1) +
  geom_step(data = df.pct %>% filter(total > 100, temp == '28C'),  aes(x = reorder(cell_type_broad, pct_both), y = pct_both, group = 1), col = 'darkgrey') +
  geom_bar(data = df.pct %>% filter(total > 100),  aes(x = reorder(cell_type_broad, pct_both), y = (pct_both_ratio)), stat = 'identity', fill = 'orange2') +
  monocle3:::monocle_theme_opts() +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab('percent cells expressing both translation and UPR') +
  xlab('') +
  ylim(-30,30) + 
  coord_polar(start = 0, clip = 'on', theta = 'x')


```


## notochord subtype markers

```{r}

plot_umap_gene_expr(cds, 'rab32a', cell_size = 0.5) + facet_grid(.~temp)
plot_umap_gene_expr(cds, 'cav3', cell_size = 0.5) + facet_grid(.~temp)
plot_umap_gene_expr(cds, 'col8a1a', cell_size = 0.5) + facet_grid(.~temp)
plot_umap_gene_expr(cds, 'entpd5a', cell_size = 0.5) + facet_grid(.~temp)


```

## hotspot umap

## read in notochord cds


## prepare combined cds for hotspot analysis

```{r}

cds <- readRDS('data/zscape/hotfish-notochord-cds.rds')
colData(cds)$embryo <- colData(cds1)$Oligo


# plot notochord by hotspot
plot_cells(cds2, color_cells_by = 'temp_up')
# plot notochord by mean nn time (timepoint)
plot_cells(cds2, color_cells_by = 'mean_nn_time')





```



## perform hotspot, plot spatially biased cells

```{r}

source("functions/hotspot_functions.R")

## hotspot label
table(colData(cds)$geno_treatment)
label_of_interest = c('injcontrol_32', 'injcontrol_34') ### set this to label of interest - here showing high temperature cells. 'ISRIB_34' 
colData(cds)$hot_label <- NA
colData(cds)$hot_label[colData(cds)$geno_treatment %in% 'injcontrol_28']  <- 'no'
colData(cds)$hot_label[colData(cds)$geno_treatment %in% label_of_interest] <- 'yes'
cds_sub = cds[,!is.na(colData(cds)$hot_label)]

hs.df = calc_hotspot(cds_sub,
                         compare_col = "hot_label", compare = "yes", 
                         reduction_method = "UMAP")
colData(cds_sub)$subset_umap2d_1 = reducedDims(cds_sub)[["UMAP"]][,1]
colData(cds_sub)$subset_umap2d_2 = reducedDims(cds_sub)[["UMAP"]][,2]
coldata_sub = colData(cds_sub) %>%
  as.data.frame() %>% 
  left_join(hs.df %>% select(cell, adjusted_pval, localG), 
            by = "cell") 

### output hotspot umap
sig_thresh = 0.05

coldata_sub$is_hotspot = NA 
coldata_sub = coldata_sub %>% 
  mutate(is_hotspot = adjusted_pval < 0.05)

# this one shows hotspot, is plotted over reference cells
ggplot() +
  geom_point(data = coldata_sub %>% filter(is.na(adjusted_pval)),
             aes(x = subset_umap2d_1,
                 y = subset_umap2d_2),
             color = "black",
             stroke = 0,
             size = 0.75) +
  geom_point(data = coldata_sub %>% filter(is.na(adjusted_pval)),
             aes(x = subset_umap2d_1,
                 y = subset_umap2d_2),
             color = "grey90",
             stroke = 0,
             size = 0.6) +
  geom_point(data = coldata_sub %>% filter(is_hotspot),
             aes(x = subset_umap2d_1,
                 y = subset_umap2d_2),
             color = 'tomato',
             stroke = 0,
             size = 0.6) +
  theme_void() +
  theme(legend.position = "none") 

