---
title: "Figure 2"
output: html_notebook
---

# read in packages

```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(plotly)
  library(viridis)
  library(data.table)
  library(DelayedArray)
  library(devtools)
  library(tidyr)
  library(monocle3)
  library(plotly)
  library(htmlwidgets)
  library(ggrepel)
  library(VGAM)
  library(pheatmap)
  library(tidymodels)
  library(stringr)
  library(gprofiler2)
  library(ggridges)
  
  DelayedArray:::set_verbose_block_processing(TRUE)
  options(DelayedArray.block.size=1e9)  
})

```


# Figure 2

## embryo trajectory

### read in embryo composition coldata

```{r}

coldat_ref = fread(file = 'data/zscape/reference-embryo-trajectory.txt'
coldat_quer = fread(file = 'data/zscape/hotfisih-query-embryo-trajectory.txt')

```

### plot embryo trajectories

```{r}

timepoint_filt = 24
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 30), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 

timepoint_filt = 30
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 30), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 

# 36hpf
timepoint_filt = 36
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 36), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 


# plot with centroids
common_cols <- intersect(names(coldat_ref), names(coldat_quer))
df1_common <- coldat_ref %>% select(common_cols)
df2_common <- coldat_quer %>% select(common_cols)
df <- bind_rows(df1_common, df2_common)
df = df %>% filter(sample == 'HF4')
df$group <- interaction(df$temp, df$timepoint)
centroids <- aggregate(cbind(umap_1, umap_2) ~ group, df, mean)
centroids = centroids %>% separate(group, into = c('temp', 'timepoint'))

# plot this out
ggplot(coldat_ref %>% filter(sample == 'HF4', fake == FALSE), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag)), aes(x = umap_1, y = umap_2), col = 'grey', cex = 3) +
  geom_point(data = coldat_quer %>% filter(temp == '32C'), aes(x = umap_1, y = umap_2), col = '#FF2500', cex = 3) +
  geom_point(data = coldat_quer %>% filter(temp == '34C'), aes(x = umap_1, y = umap_2), col = '#8A1000', cex = 3) +
  geom_point(data = centroids, aes(x = umap_1, y = umap_2, col = temp), 
                    size = 4, shape = 8) +
  scale_color_manual(values = c('grey', '#FF2500', '#8A1000')) +
  facet_grid(.~timepoint) +
  monocle3:::monocle_theme_opts() 


# plot all, including reference embryos, remove the modeled embryos 'fake'
ggplot(coldat_ref %>% filter(sample != 'HF4', fake == FALSE), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(fake == FALSE), aes(x = umap_1, y = umap_2, col = timepoint)) +
  scale_color_viridis_c() 

```


## pseudostage vs expected stage

From Kimmel et al
Hence developmental rate varies as a linear function of incubation temperature, and a simple calculation allows one to determine approximately when embryos developing at any temperature between the extremes will reach a desired stage of interest: H, = h/(0.055T - 0.57), where H, = hours of development at temperature T, and h = hours of development to reach the stage at 28.5", as set out in Table 1. For example, computation of development to the 20-somite stage (occurring at 19 h at the standard temperature) yields 23.6 hours at 25"C, and 15 hours at 33°C. These times do not differ significantly from those observed (23.5 hours and 14.5 hours).



```{r}

coldat = fread('data/zscape/hotfish-ref_18-48hpf_coldata.csv')


# compute expected stage
coldat = coldat %>% 
  mutate(expected_stage = as.numeric(substr(timepoint, 1, 2)) * (0.055 * as.numeric(substr(temp, 1, 2)) - 0.57)) 

df.emb = coldat %>% 
  select(embryo, temp, timepoint, pseudostage, expected_stage, expt) %>% 
  distinct()
df.emb$expt[grepl('HF4', df.emb$embryo)] <- 'HF4'


ggplot(df.emb %>% filter(expt %in% 'HF4'), aes(x = expected_stage, y = pseudostage, col = temp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_manual(values = c('grey','#FF6458', '#FF2003')) +
  monocle3:::monocle_theme_opts()

# another way to ask - is the avg addition in stage the same?=
se = function(x) {sd(x, na.rm=TRUE) / sqrt(length(x))}

df.stage = 
df.emb %>% 
  filter(timepoint %in% c(24,30,36)) %>% 
  group_by(temp, timepoint) %>% 
  summarize(expected = median(expected_stage),
            pseudostage.stderr = sd(pseudostage),
            pseudostage = median(pseudostage))


# pseudostage expectations
df.stage$expected_add_hr = df.stage$expected - df.stage[df.stage$temp == '28C',]$expected
df.stage$expected_add_ps = df.stage$pseudostage - df.stage[df.stage$temp == '28C',]$pseudostage
df.stage$expected_diff = df.stage$expected_add_hr - df.stage$expected_add_ps

#  scale to match hpf
df.stage$pseudostage_scaled = df.stage$pseudostage + (df.stage$expected - df.stage$pseudostage)[1:3]


# plot pseudstage and expected stage
p <- ggplot(df.stage, aes(x = expected, y = pseudostage_scaled)) +
  geom_smooth(method = "lm", se = FALSE, col = '#D5D5D5') +
  geom_abline(lty = 'longdash') + 
  geom_linerange(aes(ymin = pseudostage_scaled - pseudostage.stderr,
                     ymax = pseudostage_scaled + pseudostage.stderr), size = 0.5) +
  geom_point(aes(col = temp), size = 2) +
  scale_color_manual(values = c('grey','#FF6458', '#FF2003')) +
  monocle3:::monocle_theme_opts() +

  xlim(20, 50) + 
  ylim(20, 50) +
  theme(legend.position = "none",
          #rect = element_rect(fill = "transparent"),
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA)) 
p



# plot pseudostage against mean timepoint of neighbors

# re-shape embryo data
df.emb = coldat %>% 
  group_by(embryo) %>% 
  mutate(mean_nn_time = mean(mean_nn_time)) %>% 
  ungroup() %>% 
  select(embryo, temp, timepoint, pseudostage, expected_stage, expt, mean_nn_time) %>% 
  distinct()
df.emb$expt[grepl('HF4', df.emb$embryo)] <- 'HF4'


df.stage = 
df.emb %>% 
  group_by(temp, timepoint) %>% 
  summarize(expected = median(expected_stage),
            pseudostage.err = sd(pseudostage),
            pseudostage = median(pseudostage),
            pseudostage_rescale = pseudostage + (24 - 8.95),
            nn_time.err = sd(mean_nn_time),
            nn_time = mean(mean_nn_time))


# scatterplot mean_nn_time vs pseudostage
p <- ggplot(df.stage %>% filter( !(timepoint == 96 | timepoint == 72),
                                 temp == '28C'),
            aes(x = nn_time, y = pseudostage_rescale)) +
  geom_smooth(method = "lm", se = FALSE, col = 'black') +
  geom_linerange(aes(ymin = pseudostage_rescale - pseudostage.err,
                     ymax = pseudostage_rescale + pseudostage.err), size = 0.5, col = 'grey') +
  geom_linerange(aes(xmin = nn_time - nn_time.err,
                     xmax = nn_time + nn_time.err), size = 0.5, col = 'grey') +
  geom_point(aes(fill = timepoint), size = 3, pch = 21, col = 'black') +
  monocle3:::monocle_theme_opts() +
  scale_fill_viridis_c() +
  theme(legend.position = "none",
          #rect = element_rect(fill = "transparent"),
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA)) 

p



```


## plot cell abundance results

```{r}
celltype_disps.raw = fread('~/Dropbox/q_lab/projects/40_fish/HF4/variance/hf4_finalvarianceModelRaw_outs_hf-ctrls.csv')

# cell abundance
df.mean = fread('/Users/dorrity/Dropbox/manuscripts/stress_Dev/submission_cell/supplemental_files/Supplemental_Table_1_abundanceOuts.csv')




# heatmap of abundances
mat = 
df.mean %>% 
  filter(geno.v.wt == '34C') %>% 
  select(cell_group, timepoint, abund_log2fc) %>% 
  pivot_wider(names_from = timepoint, values_from = abund_log2fc,names_prefix = 'ps_') %>% 
  mutate(across(c(ps_1, ps_2, ps_3, ps_4), ~replace(., is.na(.), 0))) %>% 
  column_to_rownames(var = "cell_group") %>% 
  as.matrix()

# heatmap of abundances
paletteLength = 12
myColor <- colorRampPalette(c('steelblue','white', "darkred"))(paletteLength)
myBreaks <- c(seq(min(mat,na.rm=TRUE), 0, length.out=ceiling(paletteLength/2) + 1),
              seq(max(mat,na.rm=TRUE)/paletteLength, max(mat,na.rm=TRUE), length.out=floor(paletteLength/2)))
pheatmap::pheatmap(mat, col = myColor, breaks = myBreaks, cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE)



# CV heatmap
df.var = fread('data/for-zscape/hotfish-varianceModel-out.txt')

# process variance output
mat.28 = 
df.var %>% 
  filter(genotype == 'WT') %>% 
  select(cell_group, pseudostage.bin.n, cells_per_embryo_cv) %>% 
  group_by(cell_group, pseudostage.bin.n) %>% 
  summarize(avg_cv = mean(cells_per_embryo_cv, na.rm = TRUE)) %>% 
  pivot_wider(names_from = pseudostage.bin.n, values_from = avg_cv ,names_prefix = 'ps_') %>% 
  select(ps_1, ps_2, ps_3, ps_4) %>% 
  mutate(across(c(ps_1, ps_2, ps_3, ps_4), ~replace(., is.na(.), 0))) %>% 
  column_to_rownames(var = "cell_group") %>% 
  as.matrix()

mat.34 = 
df.var %>% 
  filter(genotype == '34C') %>% 
  select(cell_group, pseudostage.bin.n, cells_per_embryo_cv) %>% 
  group_by(cell_group, pseudostage.bin.n) %>% 
  summarize(avg_cv = mean(cells_per_embryo_cv, na.rm = TRUE)) %>% 
  pivot_wider(names_from = pseudostage.bin.n, values_from = avg_cv ,names_prefix = 'ps_') %>% 
  mutate(across(c(ps_1, ps_2, ps_3, ps_4), ~replace(., is.na(.), 0))) %>% 
  column_to_rownames(var = "cell_group") %>% 
  as.matrix()


# compute relative CV
common_cts = intersect(rownames(mat.28), rownames(mat.34))
mat.28 = mat.28[common_cts,]
mat.34 = mat.34[common_cts,]
mat = mat.34/mat.28
mat[mat > 3] <- 3 # set major outliers to 3 (mean + 2sd)
mat[mat < 1] <- 1 # set relative decreases to 1

# heatmap of cv
paletteLength = 12
myColor <- colorRampPalette(c('white','#f5d314', "#f56a14"))(paletteLength)
pheatmap::pheatmap(mat, col = myColor, cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = TRUE, show_colnames = TRUE)


```



## plot pheno severity on umap


```{r}

# read in abundance counts by phenotypic severity
df = fread('data/zscape/pheno-severity-abundance.txt')
coldat = fread('data/zscape//hotfish-coldata-umap.txt')

# subset to 34C, and compute log2 abundance ratio for severe vs norm
cell_df = df %>% 
  filter(genotype == '34C') %>% 
  group_by(cell_group, pheno_note) %>% 
  summarize(mean_count = mean(cells, na.rm = TRUE)) %>% 
  pivot_wider(names_from = pheno_note, values_from = mean_count) %>% 
  mutate(severe_over_norm_log2 = log2((severe/(norm + 0.01)))) %>% 
  mutate(severe_over_norm_log2 = ifelse(severe_over_norm_log2 > 2, 2, 
                                 ifelse(severe_over_norm_log2 < -2, -2, severe_over_norm_log2))) %>% 
  mutate(severe_over_norm_log2 = ifelse(norm == 0 & severe == 0, 0, severe_over_norm_log2))

# merge in cell abundance results
coldat$severe_over_norm_log2 = test$severe_over_norm_log2[match(coldat$cell_type_broad, cell_df$cell_group)]


ggplot() +
  geom_point(data = coldat %>% filter(experiment == 'HF4',
                                      !is.na(severe_over_norm_log2)),
             aes(x = umap_proj_global_1,
                 y = umap_proj_global_2),
             color = "black",
             stroke = 0,
             size = 0.4) +
  geom_point(data = coldat %>% filter(experiment == 'HF4',
                                      !is.na(severe_over_norm_log2)),
             aes(x = umap_proj_global_1,
                 y = umap_proj_global_2,
             color = severe_over_norm_log2),
             stroke = 0,
             size = 0.3) +
  scale_color_gradient2(low = 'steelblue', mid =  'white', high= 'tomato') +
  theme_void() +
  theme(legend.position = "none") 


```



#Figure S2

## molecular vs cellular staging

```{r}

coldat = fread('data/zscape/hotfish-coldata-umap.txt')

df = coldat %>% 
         group_by(embryo) %>% 
         summarize(age_ps = mean(pseudostage, na.rm=TRUE),
                   age_nn = mean(mean_nn_time, na.rm=TRUE)) %>% 
         right_join(coldat %>%
                    select(embryo, timepoint, experiment_batch, timepoint, temp) %>%
                    distinct() %>% ungroup())
df$expected_stage = as.numeric(substr(df$timepoint, 1, 2)) * (0.055 * as.numeric(substr(df$temp, 1, 2)) - 0.57)


# plot embryo stages through time, color by temp
ggplot(df %>% 
         filter(age_ps < 60)) +
  geom_point(aes(x = age_ps, y = age_nn, fill = temp), pch = 21, stroke = 1) +
  geom_smooth(aes(x = age_ps, y = age_nn, col = temp)) +
  scale_color_manual(values = c('grey', '#fa8778', '#eb4934')) +
  xlab('morphological staging (pseudostage)') +
  ylab('molecular staging (nn-transcriptional age)') +
  scale_fill_manual(values = c('grey', '#fa8778', '#eb4934')) +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none')


## CV barplots
var_df = 
df %>% 
  group_by(stage_bin, temp) %>% 
  summarize(mean_ps = mean(age_ps, na.rm=TRUE),
            var_ps = sd(age_ps, na.rm=TRUE),
            cv_ps = sd(age_ps, na.rm=TRUE)/mean(age_ps, na.rm = TRUE),
            mean_nn = mean(age_nn, na.rm=TRUE),
            var_nn = sd(age_nn, na.rm=TRUE),
            cv_nn = sd(age_nn, na.rm=TRUE)/mean(age_nn, na.rm = TRUE))

# pseudostage cv
ggplot(var_df %>% filter(temp == '28C')) +
  geom_bar(aes(x = stage_bin, y = cv_ps, fill = stage_bin),
           stat = 'identity', col = 'black') + 
  scale_fill_viridis_d() +
  ylim(0,0.32) +
  theme(text = element_text(size=8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  monocle3:::monocle_theme_opts() +
  ylab('CV (pseudostage)') + xlab('timepoint bin (hpf)') +
  theme(legend.position = 'none',
        axis.line.x.bottom = element_line(colour = 'black', size = 0.7),
        axis.line.y.left = element_line(colour = 'black', size = 0.7)) 

# mean nn time cv
ggplot(var_df %>% filter(temp == '28C')) +
  geom_bar(aes(x = stage_bin, y = cv_nn, fill = stage_bin),
           stat = 'identity', col = 'black') + 
  scale_fill_viridis_d() +
  ylim(0,0.32) +
  theme(text = element_text(size=8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  monocle3:::monocle_theme_opts() +
  ylab('CV (nn age)') + xlab('timepoint bin (hpf)') +
  theme(legend.position = 'none',
        axis.line.x.bottom = element_line(colour = 'black', size = 0.7),
        axis.line.y.left = element_line(colour = 'black', size = 0.7)) 


```

