---
title: "Figure 2"
output: html_notebook
---

# read in packages

```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(plotly)
  library(viridis)
  library(data.table)
  library(DelayedArray)
  library(devtools)
  library(tidyr)
  library(monocle3)
  library(plotly)
  library(htmlwidgets)
  library(ggrepel)
  library(VGAM)
  library(pheatmap)
  library(tidymodels)
  library(stringr)
  library(gprofiler2)
  library(ggridges)
  
  DelayedArray:::set_verbose_block_processing(TRUE)
  options(DelayedArray.block.size=1e9)  
})

```


# Figure 2

## embryo trajectory

### read in embryo composition coldata

```{r}

coldat_ref = fread(file = 'data/zscape/reference-embryo-trajectory.txt'
coldat_quer = fread(file = 'data/zscape/hotfisih-query-embryo-trajectory.txt')

```

### plot embryo trajectories

```{r}

timepoint_filt = 24
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 30), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 

timepoint_filt = 30
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 30), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 

# 36hpf
timepoint_filt = 36
ggplot(coldat_ref %>% filter(sample == 'HF4', timepoint == 36), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag), , timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = 'grey') +
  geom_point(data = coldat_quer %>% filter(temp == '32C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#FF2500') +
  geom_point(data = coldat_quer %>% filter(temp == '34C', timepoint == timepoint_filt), aes(x = umap_1, y = umap_2), col = '#8A1000') +
  scale_color_viridis_c() 


# plot with centroids
common_cols <- intersect(names(coldat_ref), names(coldat_quer))
df1_common <- coldat_ref %>% select(common_cols)
df2_common <- coldat_quer %>% select(common_cols)
df <- bind_rows(df1_common, df2_common)
df = df %>% filter(sample == 'HF4')
df$group <- interaction(df$temp, df$timepoint)
centroids <- aggregate(cbind(umap_1, umap_2) ~ group, df, mean)
centroids = centroids %>% separate(group, into = c('temp', 'timepoint'))

# plot this out
ggplot(coldat_ref %>% filter(sample == 'HF4', fake == FALSE), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  geom_point(data = coldat_ref %>% filter(temp == '28C', grepl('HF4', oligo_tag)), aes(x = umap_1, y = umap_2), col = 'grey', cex = 3) +
  geom_point(data = coldat_quer %>% filter(temp == '32C'), aes(x = umap_1, y = umap_2), col = '#FF2500', cex = 3) +
  geom_point(data = coldat_quer %>% filter(temp == '34C'), aes(x = umap_1, y = umap_2), col = '#8A1000', cex = 3) +
  geom_point(data = centroids, aes(x = umap_1, y = umap_2, col = temp), 
                    size = 4, shape = 8) +
  scale_color_manual(values = c('grey', '#FF2500', '#8A1000')) +
  facet_grid(.~timepoint) +
  monocle3:::monocle_theme_opts() 


# plot all, including reference embryos, remove the modeled embryos 'fake'
ggplot(coldat_ref %>% filter(sample != 'HF4', fake == FALSE), aes(x = umap_1, y = umap_2)) + 
  geom_point(col = 'grey') +
  #geom_path(data = coldat_ref[coldat_ref$sample != 'HF4',], aes(x = umap_1, y = umap_2), col = 'black') +
  geom_path(data = coldat_ref[coldat_ref$fake == TRUE,], aes( x = umap_1, y = umap_2, group = 1), col = 'lightgrey', size = 2) +
  geom_point(data = coldat_ref %>% filter(fake == FALSE), aes(x = umap_1, y = umap_2, col = timepoint)) +
  scale_color_viridis_c() 

```


## pseudostage vs expected stage

From Kimmel et al
Hence developmental rate varies as a linear function of incubation temperature, and a simple calculation allows one to determine approximately when embryos developing at any temperature between the extremes will reach a desired stage of interest: H, = h/(0.055T - 0.57), where H, = hours of development at temperature T, and h = hours of development to reach the stage at 28.5", as set out in Table 1. For example, computation of development to the 20-somite stage (occurring at 19 h at the standard temperature) yields 23.6 hours at 25"C, and 15 hours at 33°C. These times do not differ significantly from those observed (23.5 hours and 14.5 hours).



```{r}

coldat = fread('data/zscape/hotfish-ref_18-48hpf_coldata.csv')


# compute expected stage
coldat = coldat %>% 
  mutate(expected_stage = as.numeric(substr(timepoint, 1, 2)) * (0.055 * as.numeric(substr(temp, 1, 2)) - 0.57)) 

df.emb = coldat %>% 
  select(embryo, temp, timepoint, pseudostage, expected_stage, expt) %>% 
  distinct()
df.emb$expt[grepl('HF4', df.emb$embryo)] <- 'HF4'


ggplot(df.emb %>% filter(expt %in% 'HF4'), aes(x = expected_stage, y = pseudostage, col = temp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_manual(values = c('grey','#FF6458', '#FF2003')) +
  monocle3:::monocle_theme_opts()

# another way to ask - is the avg addition in stage the same?=
se = function(x) {sd(x, na.rm=TRUE) / sqrt(length(x))}

df.stage = 
df.emb %>% 
  filter(timepoint %in% c(24,30,36)) %>% 
  group_by(temp, timepoint) %>% 
  summarize(expected = median(expected_stage),
            pseudostage.stderr = sd(pseudostage),
            pseudostage = median(pseudostage))


# pseudostage expectations
df.stage$expected_add_hr = df.stage$expected - df.stage[df.stage$temp == '28C',]$expected
df.stage$expected_add_ps = df.stage$pseudostage - df.stage[df.stage$temp == '28C',]$pseudostage
df.stage$expected_diff = df.stage$expected_add_hr - df.stage$expected_add_ps

#  scale to match hpf
df.stage$pseudostage_scaled = df.stage$pseudostage + (df.stage$expected - df.stage$pseudostage)[1:3]


# plot pseudstage and expected stage
p <- ggplot(df.stage, aes(x = expected, y = pseudostage_scaled)) +
  geom_smooth(method = "lm", se = FALSE, col = '#D5D5D5') +
  geom_abline(lty = 'longdash') + 
  geom_linerange(aes(ymin = pseudostage_scaled - pseudostage.stderr,
                     ymax = pseudostage_scaled + pseudostage.stderr), size = 0.5) +
  geom_point(aes(col = temp), size = 2) +
  scale_color_manual(values = c('grey','#FF6458', '#FF2003')) +
  monocle3:::monocle_theme_opts() +

  xlim(20, 50) + 
  ylim(20, 50) +
  theme(legend.position = "none",
          #rect = element_rect(fill = "transparent"),
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA)) 
p



# plot pseudostage against mean timepoint of neighbors

# re-shape embryo data
df.emb = coldat %>% 
  group_by(embryo) %>% 
  mutate(mean_nn_time = mean(mean_nn_time)) %>% 
  ungroup() %>% 
  select(embryo, temp, timepoint, pseudostage, expected_stage, expt, mean_nn_time) %>% 
  distinct()
df.emb$expt[grepl('HF4', df.emb$embryo)] <- 'HF4'


df.stage = 
df.emb %>% 
  group_by(temp, timepoint) %>% 
  summarize(expected = median(expected_stage),
            pseudostage.err = sd(pseudostage),
            pseudostage = median(pseudostage),
            pseudostage_rescale = pseudostage + (24 - 8.95),
            nn_time.err = sd(mean_nn_time),
            nn_time = mean(mean_nn_time))


# scatterplot mean_nn_time vs pseudostage
p <- ggplot(df.stage %>% filter( !(timepoint == 96 | timepoint == 72),
                                 temp == '28C'),
            aes(x = nn_time, y = pseudostage_rescale)) +
  geom_smooth(method = "lm", se = FALSE, col = 'black') +
  geom_linerange(aes(ymin = pseudostage_rescale - pseudostage.err,
                     ymax = pseudostage_rescale + pseudostage.err), size = 0.5, col = 'grey') +
  geom_linerange(aes(xmin = nn_time - nn_time.err,
                     xmax = nn_time + nn_time.err), size = 0.5, col = 'grey') +
  geom_point(aes(fill = timepoint), size = 3, pch = 21, col = 'black') +
  monocle3:::monocle_theme_opts() +
  scale_fill_viridis_c() +
  theme(legend.position = "none",
          #rect = element_rect(fill = "transparent"),
          panel.background = element_rect(fill = "transparent",colour = NA),
          plot.background = element_rect(fill = "transparent",colour = NA)) 

p



```

### pseudostage vs mean_nn_time stage


```{r}

se = function(x) {sd(x, na.rm=TRUE)/sqrt(sum(x, na.rm=TRUE))}

df = coldat %>% 
  group_by(oligo_tag, timepoint_fix, temp) %>% 
  dplyr::summarize(mean_nn = mean(mean_nn_time, na.rm=TRUE),
            mean_ps = mean(pseudostage, na.rm=TRUE),
            mean_nn_time.err = se(mean_nn_time),
            pseudostage.err = se(pseudostage))


df.g = df %>% 
  group_by(timepoint_fix, temp) %>% 
  summarize(group.mean_nn = mean(mean_nn, na.rm=TRUE),
            group.pseudostage = mean(mean_ps, na.rm=TRUE),
            group.mean_nn.err = sd(mean_nn),
            group.pseudostage.err = sd(mean_ps))

```


### scatterplots 

```{r}

timecols = c("#DF4828", "#E78C35", "#F6C141",  "#4EB265",  "#1965B0")
time_pal = colorRampPalette(timecols)(length(unique(df.g$timepoint_fix))) 
temp_pal = c('grey','#FF6458', '#FF2003') 

# show 28C only
ggplot(df.g %>% filter(temp == '28C'), aes(x = group.mean_nn, y = group.pseudostage)) +
  geom_smooth(method = 'lm', se = FALSE, col = 'black', size = 0.5) + 
  geom_linerange(aes(ymin = group.pseudostage - group.pseudostage.err,
                     ymax = group.pseudostage + group.pseudostage.err), size = 0.5, col = 'grey') +
  geom_linerange(aes(xmin = group.mean_nn - group.mean_nn.err,
                     xmax = group.mean_nn + group.mean_nn.err), size = 0.5, col = 'grey') +
  geom_point(aes(fill = factor(timepoint_fix)), pch = 21, col = 'black', size = 2) +
  scale_fill_manual(values = time_pal) +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none') + 
  ylab('pseudostage') + xlab('average timepoint of nearest-neighbors in reference')

cor(df.g$group.mean_nn, df.g$group.pseudostage, method = 'pearson')

# show all
ggplot(df.g, aes(x = group.mean_nn, y = group.pseudostage)) +
  geom_smooth(method = 'lm', se = FALSE, col = 'black', size = 0.5) + 
  geom_linerange(aes(ymin = group.pseudostage - group.pseudostage.err,
                     ymax = group.pseudostage + group.pseudostage.err), size = 0.5, col = 'grey') +
  geom_linerange(aes(xmin = group.mean_nn - group.mean_nn.err,
                     xmax = group.mean_nn + group.mean_nn.err), size = 0.5, col = 'grey') +
  geom_point(aes(fill = factor(timepoint_fix)), pch = 21, col = 'black', size = 2) +
  scale_fill_manual(values = time_pal) +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none') + 
  ylab('pseudostage') + xlab('average timepoint of nearest-neighbors in reference')

cor(df.g$group.mean_nn, df.g$group.pseudostage, method = 'pearson')



# show HF timepoints only

ggplot(df.g %>% filter(timepoint_fix %in% c(24,30,36)), aes(x = group.mean_nn, y = group.pseudostage)) +
  # geom_smooth(method = 'lm', se = FALSE, col = 'black', size = 0.5) + 
  geom_linerange(aes(ymin = group.pseudostage - group.pseudostage.err,
                     ymax = group.pseudostage + group.pseudostage.err), size = 0.5, col = 'grey') +
  geom_linerange(aes(xmin = group.mean_nn - group.mean_nn.err,
                     xmax = group.mean_nn + group.mean_nn.err), size = 0.5, col = 'grey') +
  geom_point(aes(fill = temp), pch = 21, col = 'black', size = 2) +
  scale_fill_manual(values = temp_pal) +
  monocle3:::monocle_theme_opts() +
  facet_wrap(. ~ timepoint_fix) +
  ylab('pseudostage') + xlab('average timepoint of nearest-neighbors in reference')

cor(df.g$group.mean_nn, df.g$group.pseudostage, method = 'pearson')

```



## plot changes in cell abudance + variance

### read in model outs

```{r}

means = fread('data/zscape/hotfish-mean-model-outs.txt')
cvs = fread('data/zscape/hotfish-cv-model-outs.txt')

# make sure they share cell types
common_cts = intersect(unique(means$cell_group), unique(cvs$cell_group))
means = means %>% filter(cell_group %in% common_cts)
cvs = cvs %>% filter(cell_group %in% common_cts)

```

### plot heatmaps - mean

```{r}

# plot mean abundance heatmap with significance boxes
res_df = means
res_df$timepoint = res_df$timepoint.bin.n

geno = '34C'
sig_thresh = 5e-2

sig_df = res_df %>% 
    ungroup() %>% 
    filter(geno.v.wt == geno) %>%
    filter(cell_group != "NA") %>%
    dplyr::mutate(sig_fc = case_when(qval < sig_thresh ~ TRUE,
                                     TRUE ~ FALSE))

# get rid of cell types with no significant changes
keep = sig_df %>% 
    filter(sig_fc) %>% 
    distinct(cell_group) %>% pull %>% 
    as.character

hm_wide = sig_df %>%
    group_by(timepoint.bin.n, cell_group) %>%
      summarize(abund_log2fc = mean(abund_log2fc, na.rm=TRUE)) %>%
      dplyr::rename(timepoint = timepoint.bin.n) %>% 
    select(timepoint, cell_group, abund_log2fc) %>%
    pivot_wider(names_from = cell_group,
                values_from = abund_log2fc, values_fill = c(0))

hm_plot = hm_wide %>% 
    pivot_longer(!timepoint, names_to = "cell_group", values_to = "plot") %>%
    left_join(sig_df %>%
                select(timepoint, cell_group, sig_fc), by = c("timepoint", "cell_group")) %>%
    mutate(timepoint = as.factor(timepoint),
           cell_group = as.factor(cell_group))

# get clustering results
hm_mat <- as.matrix(hm_wide[, -1])
rownames(hm_mat) <- hm_wide$timepoint
hm_mat <- t(hm_mat)
hm.dendro <- as.dendrogram(hclust(d = dist(x = hm_mat)))
hm.order <- order.dendrogram(hm.dendro)
# reorder cell type factor
hm_plot$cell_group <- factor(as.character(hm_plot$cell_group),
                               rownames(hm_mat[hm.order,]))

# make df for plotting significance boxes
fms = hm_plot[hm_plot$sig_fc, c("timepoint", "cell_group")]
fms$timepoint = as.integer(fms$timepoint)
fms$cell_group = as.integer(fms$cell_group)

ggplot(hm_plot, 
       aes(x = timepoint, y = cell_group, fill = plot)) + 
  geom_tile() +
  scale_fill_gradient2(low = "#3234a9", mid = "white", high= "#7B0017",
                       na.value="black",  limits = c(-2.0, 2.0), midpoint = 0) +
  geom_rect(data=fms, size=0.5, fill=NA, colour="black",
            aes(xmin=timepoint - 0.5, xmax=timepoint + 0.5, ymin=cell_group - 0.5, ymax=cell_group + 0.5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_blank(), text = element_text(size = 12),
        legend.title = element_blank(), plot.title = element_text(hjust = 0.5),
        plot.margin=unit(c(.5,.5,.5,2.5),"cm"), legend.key.size = unit(0.5, 'cm'),
        axis.line = element_line(colour = "black", size = 1), 
        axis.ticks = element_line(colour = "black", size = 0.25)) +
  ggtitle(geno) +
  labs(fill = "mean log2FC") +
  xlab('pseudostage bin') + 
  coord_equal() +
  monocle3:::monocle_theme_opts() 


# grab ordering for cv plot
ct_ord = levels(hm_plot$cell_group)


```

### plot heatmaps - cv

```{r}

res_df = cvs
res_df$timepoint = res_df$timepoint.bin.n

geno = '34C'
sig_thresh = 5e-2

sig_df = res_df %>% 
    ungroup() %>% 
    filter(genotype == geno) %>%
    filter(cell_group != "NA") %>%
    filter(cv_rel != 0) %>% 
    dplyr::mutate(sig_fc = case_when(cv_z_stat_p_val < sig_thresh ~ TRUE,
                                     TRUE ~ FALSE))

# get rid of cell types with no significant changes
keep = sig_df %>% 
    filter(sig_fc) %>% 
    distinct(cell_group) %>% pull %>% 
    as.character

hm_wide = sig_df %>%
    group_by(timepoint.bin.n, cell_group) %>%
      summarize(cv_rel = mean(cv_rel, na.rm=TRUE)) %>%
      dplyr::rename(timepoint = timepoint.bin.n) %>% 
    select(timepoint, cell_group, cv_rel) %>%
    pivot_wider(names_from = cell_group,
                values_from = cv_rel, values_fill = c(0))

hm_plot = hm_wide %>% 
    pivot_longer(!timepoint, names_to = "cell_group", values_to = "plot") %>%
    left_join(sig_df %>%
                select(timepoint, cell_group, sig_fc), by = c("timepoint", "cell_group")) %>%
    mutate(timepoint = as.factor(timepoint),
           cell_group = as.factor(cell_group))

# reorder cell type factor by mean results
hm_plot$cell_group <- factor(as.character(hm_plot$cell_group),
                               ct_ord)

# make df for plotting significance boxes
fms = hm_plot[hm_plot$sig_fc, c("timepoint", "cell_group")]
fms$timepoint = as.integer(fms$timepoint)
fms$cell_group = as.integer(fms$cell_group)

# show non-log relative cv, set color palette
colpal = '#fef6b5,#ffdd9a,#ffc285,#ffa679,#fa8a76,#f16d7a,#e15383'
var_colpal = unlist(strsplit(colpal, split = ','))

# # set relative decreases to 1
hm_plot$plot[hm_plot$plot < 1] <- NA

# set major outliers to 3 (mean + 2sd)
cv_top = 3.0
hm_plot$plot[hm_plot$plot >= cv_top] <- cv_top

ggplot(hm_plot, 
         aes(x = timepoint, y = cell_group, fill = plot)) + 
    geom_tile() +
    scale_fill_gradientn(colours = var_colpal, na.value = 'white') + 
    geom_rect(data=fms, size=0.5, fill=NA, colour="black",
              aes(xmin=timepoint - 0.5, xmax=timepoint + 0.5, ymin=cell_group - 0.5, ymax=cell_group + 0.5)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title.y = element_blank(), text = element_text(size = 12),
          legend.title = element_blank(), plot.title = element_text(hjust = 0.5),
          plot.margin=unit(c(.5,.5,.5,2.5),"cm"), legend.key.size = unit(0.5, 'cm'),
          axis.line = element_line(colour = "black", size = 1), 
          axis.ticks = element_line(colour = "black", size = 0.25)) +
    ggtitle(geno) +
    labs(fill = "CV log2FC") +
    xlab('pseudostage bin') + 
    coord_equal() +
    monocle3:::monocle_theme_opts()

```


## plot pheno severity on umap

```{r}

# read in abundance counts by phenotypic severity
df = fread('data/zscape/pheno-severity-abundance.txt')
coldat = fread('data/zscape//hotfish-coldata-umap.txt')

# subset to 34C, and compute log2 abundance ratio for severe vs norm
cell_df = df %>% 
  filter(genotype == '34C') %>% 
  group_by(cell_group, pheno_note) %>% 
  summarize(mean_count = mean(cells, na.rm = TRUE)) %>% 
  pivot_wider(names_from = pheno_note, values_from = mean_count) %>% 
  mutate(severe_over_norm_log2 = log2((severe/(norm + 0.01)))) %>% 
  mutate(severe_over_norm_log2 = ifelse(severe_over_norm_log2 > 2, 2, 
                                 ifelse(severe_over_norm_log2 < -2, -2, severe_over_norm_log2))) %>% 
  mutate(severe_over_norm_log2 = ifelse(norm == 0 & severe == 0, 0, severe_over_norm_log2))

# merge in cell abundance results
coldat$severe_over_norm_log2 = test$severe_over_norm_log2[match(coldat$cell_type_broad, cell_df$cell_group)]


ggplot() +
  geom_point(data = coldat %>% filter(experiment == 'HF4',
                                      !is.na(severe_over_norm_log2)),
             aes(x = umap_proj_global_1,
                 y = umap_proj_global_2),
             color = "black",
             stroke = 0,
             size = 0.4) +
  geom_point(data = coldat %>% filter(experiment == 'HF4',
                                      !is.na(severe_over_norm_log2)),
             aes(x = umap_proj_global_1,
                 y = umap_proj_global_2,
             color = severe_over_norm_log2),
             stroke = 0,
             size = 0.3) +
  scale_color_gradient2(low = 'steelblue', mid =  'white', high= 'tomato') +
  theme_void() +
  theme(legend.position = "none") 


```



#Figure S2

## molecular vs cellular staging

```{r}

coldat = fread('data/zscape/hotfish-coldata-umap.txt')

df = coldat %>% 
         group_by(embryo) %>% 
         summarize(age_ps = mean(pseudostage, na.rm=TRUE),
                   age_nn = mean(mean_nn_time, na.rm=TRUE)) %>% 
         right_join(coldat %>%
                    select(embryo, timepoint, experiment_batch, timepoint, temp) %>%
                    distinct() %>% ungroup())
df$expected_stage = as.numeric(substr(df$timepoint, 1, 2)) * (0.055 * as.numeric(substr(df$temp, 1, 2)) - 0.57)


# plot embryo stages through time, color by temp
ggplot(df %>% 
         filter(age_ps < 60)) +
  geom_point(aes(x = age_ps, y = age_nn, fill = temp), pch = 21, stroke = 1) +
  geom_smooth(aes(x = age_ps, y = age_nn, col = temp)) +
  scale_color_manual(values = c('grey', '#fa8778', '#eb4934')) +
  xlab('morphological staging (pseudostage)') +
  ylab('molecular staging (nn-transcriptional age)') +
  scale_fill_manual(values = c('grey', '#fa8778', '#eb4934')) +
  monocle3:::monocle_theme_opts() +
  theme(legend.position = 'none')


## CV barplots
var_df = 
df %>% 
  group_by(stage_bin, temp) %>% 
  summarize(mean_ps = mean(age_ps, na.rm=TRUE),
            var_ps = sd(age_ps, na.rm=TRUE),
            cv_ps = sd(age_ps, na.rm=TRUE)/mean(age_ps, na.rm = TRUE),
            mean_nn = mean(age_nn, na.rm=TRUE),
            var_nn = sd(age_nn, na.rm=TRUE),
            cv_nn = sd(age_nn, na.rm=TRUE)/mean(age_nn, na.rm = TRUE))

# pseudostage cv
ggplot(var_df %>% filter(temp == '28C')) +
  geom_bar(aes(x = stage_bin, y = cv_ps, fill = stage_bin),
           stat = 'identity', col = 'black') + 
  scale_fill_viridis_d() +
  ylim(0,0.32) +
  theme(text = element_text(size=8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  monocle3:::monocle_theme_opts() +
  ylab('CV (pseudostage)') + xlab('timepoint bin (hpf)') +
  theme(legend.position = 'none',
        axis.line.x.bottom = element_line(colour = 'black', size = 0.7),
        axis.line.y.left = element_line(colour = 'black', size = 0.7)) 

# mean nn time cv
ggplot(var_df %>% filter(temp == '28C')) +
  geom_bar(aes(x = stage_bin, y = cv_nn, fill = stage_bin),
           stat = 'identity', col = 'black') + 
  scale_fill_viridis_d() +
  ylim(0,0.32) +
  theme(text = element_text(size=8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  monocle3:::monocle_theme_opts() +
  ylab('CV (nn age)') + xlab('timepoint bin (hpf)') +
  theme(legend.position = 'none',
        axis.line.x.bottom = element_line(colour = 'black', size = 0.7),
        axis.line.y.left = element_line(colour = 'black', size = 0.7)) 


```

